---
title: "0556078_hwk03"
author: "0556078"
date: "November 18, 2016"
output: html_document
---

```{r}
library(rafalib)
library(dplyr)
```

## Question 1:
### 1-1.
**Ans: t = 0.3008**
```{r}
set.seed(1)
X <- rnorm(5)
tstat <- mean(X)*sqrt(5)/sd(X)
round(tstat, 4)
```

### 1-2.
**Ans: 0.068**
```{r}
t_sampling <- function(n){
  X <- rnorm(n)
  tstat <- mean(X)*sqrt(n)/sd(X)
  return(tstat)
}
B <- 1000
set.seed(1)
t_sample <- replicate(B, t_sampling(5))
```
```{r}
round(mean(t_sample > 2), 4)
```

### 1-3.
**Ans: C) The approximations are spot on for all sample sizes.**
```{r}
smp_size = c(5, 10, 20, 30)
mypar(2,2)
for(s in smp_size){
  set.seed(1)
  t_sample <- replicate(B, t_sampling(s))
  #ps <- (seq(0,B-1)+0.5)/B
  ps <- seq(1/(B+1), 1-1/(B+1),len=B)
  qqplot(qt(ps, df = s-1), t_sample, main=paste("degrees of freedom = ", s-1), 
         xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
  abline(0, 1, col = 2)
}
```

### 1-4.
**Ans: C) The approximations are spot on for all sample sizes.**
```{r}
t2_sampling <- function(n){
  X1 <- rnorm(n)
  X2 <- rnorm(n)
  tstat <- (mean(X1) - mean(X2))*sqrt(n)/sqrt(var(X1)+var(X2))
  return(tstat)
}
mypar(2,2)
for(s in smp_size){
  set.seed(1)
  t_sample <- replicate(B, t2_sampling(s))
  #ps <- (seq(0,B-1)+0.5)/B
  ps <- seq(1/(B+1), 1-1/(B+1),len=B)
  qqplot(qt(ps, df = 2*s-2), t_sample, main=paste("degrees of freedom = ", 2*s-2), 
         xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
  abline(0, 1, col = 2)
}
```

### 1-5.
**Ans: No.**
```{r}
mypar(1,1)
gen_tstat_binom <- function(n){
  X <- rbinom(n, size = 1, prob = 0.5)
  tstat <- sqrt(n)*mean(X)/sd(X)
  return(tstat)
}
set.seed(1)
t_sample <- replicate(B, gen_tstat_binom(15))
ps <- (seq(0,B-1)+0.5)/B
qqplot(qt(ps, df = 15-1), t_sample, main=paste("degrees of freedom = ", 15-1), 
         xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
qqline(t_sample, col = 2)
```

### 1-6.
**Ans: Yes.**
```{r}
set.seed(1)
t_sample <- replicate(B, gen_tstat_binom(500))
ps <- (seq(0,B-1)+0.5)/B
qqplot(qt(ps, df = 500-1), t_sample, main=paste("degrees of freedom = ", 500-1), 
         xlab = "Theoretical Quantiles", ylab = "Sample Quantiles")
qqline(t_sample, col = 2)
```

### 1-7.
**Ans: D) The sample median is approximately normal with mean 0 and SD larger than $1/\sqrt{N}$**  
Because the distribution of sample median is approximate standard normal distribution by observation on the following experiment.
```{r}
gen_median <- function(n){
  X <- rnorm(n)
  return(median(X))
}
set.seed(1)
m_smp_small <- replicate(B, gen_median(10))
m_smp_large <- replicate(B, gen_median(100))
```

**check: If Sample Median ~ Norm(0,1/sqrt(N))**
```{r}
mypar(1,2)
qqplot(qnorm(ps, mean = 0, sd = 1/sqrt(10)), m_smp_small, main=paste("Small Sample"),  
       xlab = "Norm Quantiles", ylab = "Small Sample Quantiles")
abline(0, 1, col = 2)
qqplot(qnorm(ps, mean = 0, sd = 1/sqrt(100)), m_smp_large, main=paste("Large Sample"),  
       xlab = "Norm Quantiles", ylab = "Large Sample Quantiles")
abline(0, 1, col = 2)
```

**check: If Sample Median ~ Norm(0,1)**
```{r}
mypar(1,2)
qqnorm(m_smp_small, main=paste("Small Sample"),  
       xlab = "Standard Norm Quantiles", ylab = "Small Sample Quantiles")
qqline(m_smp_small, col = 2)
qqnorm(m_smp_large, main=paste("Large Sample"),  
       xlab = "Standard Norm Quantiles", ylab = "Large Sample Quantiles")
qqline(m_smp_large, col = 2)
```

**check: If Sample Median ~ T**
```{r}
mypar(1,2)
qqplot(qt(ps, df = 10-1),m_smp_small, main=paste("Small Sample"),  
       xlab = "T Quantiles (df = 9)", ylab = "Small Sample Quantiles")
abline(0, 1, col = 2)
qqplot(qt(ps, df = 100-1),m_smp_large, main=paste("Large Sample"),  
       xlab = "T Quantiles (df = 99)", ylab = "Large Sample Quantiles")
abline(0, 1, col = 2)
```

## Question 2:
```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/babies.txt"
filename <- basename(url)
#download.file(url, destfile = filename)
babies <- read.table("babies.txt", header = TRUE)
bwt.nonsmoke <- filter(babies, smoke == 0) %>% select(bwt) %>% unlist
bwt.smoke <- filter(babies, smoke == 1) %>% select(bwt) %>% unlist
```

```{r}
N <- 10
set.seed(1)
nonsmokers <- sample(bwt.nonsmoke , N)
smokers <- sample(bwt.smoke , N)
obs <- mean(smokers) - mean(nonsmokers)
```

```{r}
dat <- c(smokers,nonsmokers)
shuffle <- sample( dat )
smokersstar <- shuffle[1:N]
nonsmokersstar <- shuffle[(N+1):(2*N)]
mean(smokersstar) - mean(nonsmokersstar)
```

### 2-1.
**Ans: p-value = 0.052**

**Hypotheses**  
We are interested in testing the following hypotheses:
$$ H_{0}:\mu_{x} = \mu_{y} \\
   H_{1}:\mu_{x} \neq \mu_{y} $$
where $\mu_{x}$ and $\mu_{y}$ are the population mean of smoking and non-smoking groups, respectively.

```{r}
gen_diff_mean_permut <- function(){
  shuffle <- sample(dat)
  smokersstar <- shuffle[1:N]
  nonsmokersstar <- shuffle[(N+1):(2*N)]
  return(mean(smokersstar) - mean(nonsmokersstar))
}
set.seed(1)
X_permut <- replicate(1000, gen_diff_mean_permut())
```

```{r}
p_val <- mean(abs(X_permut) > abs(obs))
round(p_val, 4)
```

### 2-2. Ans: p-value = 0.017
**Hypotheses**  
We are interested in testing the following hypotheses:
$$ H_{0}:m_{x} = m_{y} \\
   H_{1}:m_{x} \neq m_{y} $$
where $m_{x}$ and $m_{y}$ are the population median of smoking and non-smoking groups, respectively.
```{r}
obs <- median(smokers) - median(nonsmokers)

gen_diff_median_permut <- function(){
  shuffle <- sample(dat)
  smokersstar <- shuffle[1:N]
  nonsmokersstar <- shuffle[(N+1):(2*N)]
  return(median(smokersstar) - median(nonsmokersstar))
}
set.seed(1)
X_permut <- replicate(1000, gen_diff_median_permut())
```

```{r}
p_val <- mean(abs(X_permut) > abs(obs))
round(p_val, 4)
```

## Question 3:
### 3-1.
**Ans: 5.9632**
```{r}
obs <- sd(smokers) - sd(nonsmokers)

gen_diff_sd_bootstrap <- function(){
  smokersstar <- sample(smokers, N, replace = TRUE)
  nonsmokersstar <- sample(nonsmokers, N, replace = TRUE)
  return(sd(smokersstar) - sd(nonsmokersstar))
}
set.seed(1)
X_bootstrap <- replicate(1000, gen_diff_sd_bootstrap())
```

```{r}
round(sd(X_bootstrap), 4)
```

```{r}
round((obs - mean(X_bootstrap))/sd(X_bootstrap), 4)
```