---
title: "0556082_hwk03"
author: "0556082"
date: "11/20/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("rafalib")
#install.packages("dplyr")
library(dplyr)
library(rafalib)
```

## 1.1

ans: 0.3008
```{r}
set.seed(1)
generate_t <- function(n){
    sample <- rnorm(n)
  return (sqrt(n) * mean(sample) / sd(sample))

}
round(generate_t(5), 4)
```

## 1.2

ans: 0.068
```{r}
set.seed(1)
B <- 1000
sample_t <- replicate(B, generate_t(5))
round(mean(sample_t > 2), 4)
```

## 1.3

ans: (C) The approximations are spot on for all sample sizes.
```{r}
N <- c(5, 10, 20, 50)
B <- 100
mypar(2,2)
for(n in N){
    ps <- seq(1/(B+1), 1-1/(B+1),len=B)
  sample_t <- replicate(B, generate_t(n))
    title <- paste0("N=", n)
    qqplot(qt(ps, df=n-1), sample_t, main=title)
      qqline(sample_t, col = 2)

}
```

## 1.4

ans: (C) The approximations are spot on for all sample sizes.
```{r}
generate_two_nor <- function(n){
    x1 <- rnorm(n, 0, 1)
  x2 <- rnorm(n, 0, 1)
    tstat <- (mean(x1)-mean(x2)) /sqrt( var(x1)/n + var(x2)/n  ) 
    return(tstat)

}
mypar(1, 2)
B <- 100
ps <- seq(1/(B+1), 1-1/(B+1),len=B)
n <- 10
small_x <- replicate(1000, generate_two_nor(n))
title <- paste0("Small sample N=", n)
qqplot(qt(ps, df = 2*n-2), small_x, main=title)
abline(0, 1, col=2)

n <- 1000
large_x <- replicate(1000, generate_two_nor(n))
title <- paste0("Large sample N=", n)
qqplot(qt(ps, df = 2*n-2), large_x, main=title)
abline(0, 1, col=2)
```

## 1.5

ans: no
```{r}
generate_t_bi <- function(n){
    X <- rbinom(n,size=1,prob=0.5)
  tstat <- sqrt(n)*mean(X) / sd(X)
    return (tstat)

}
n <- 15
sample_t_bi <- replicate(1000, generate_t_bi(n))
title <- paste0("N=", n)
qqplot(qt(ps, df=n-1), sample_t_bi, main=title)
qqline(sample_t_bi, col = 2)
```

## 1.6

ans: yes
```{r}
n <- 500
sample_t_bi <- replicate(1000, generate_t_bi(n))
title <- paste0("N=", n)
qqplot(qt(ps, df=n-1), sample_t_bi, main=title)
qqline(sample_t_bi, col = 2)
```

## 1.7

ans: (C)The sample median is t-distributed for small samples and normally distributed for large ones.
```{r}
generate_sample_median <- function(n){
    sample_norm <- rnorm(n)
  return (median(sample_norm))

}
set.seed(1)
B <- 100
ps <- seq(1/(B+1), 1-1/(B+1),len=B)
small_sample_median <- replicate(1000, generate_sample_median(10))
large_sample_median <- replicate(1000, generate_sample_median(1000))
```

### Compare with normal with mean 0 and SD 1/sqrt(N)
```{r}
mypar(1,2)
title <- paste0("small")
qqplot(qnorm(ps, 0, 1/sqrt(10)), small_sample_median, main=title)
qqline(small_sample_median, col = 2)
title <- paste0("large")
qqplot(qnorm(ps, 0, 1/sqrt(1000)), large_sample_median, main=title)
qqline(large_sample_median, col = 2)
```

### Compare with normal with mean 0 and SD 1
```{r}
mypar(1,2)
title <- paste0("small")
qqplot(qnorm(ps, 0, 1), small_sample_median, main=title)
qqline(small_sample_median, col = 2)
title <- paste0("large")
qqplot(qnorm(ps, 0, 1), large_sample_median, main=title)
qqline(large_sample_median, col = 2)
```

### Compare with t distribution
```{r}
mypar(1,2)
title <- paste0("small")
qqplot(qt(ps, df=9), small_sample_median, main=title)
qqline(small_sample_median, col = 2)
title <- paste0("large")
qqplot(qt(ps, df=999), large_sample_median, main=title)
qqline(large_sample_median, col= 2)
```

### Compare with normal with mean 0 and SD large than 1/sqrt(N)
```{r}
mypar(1,2)
title <- paste0("small")
qqplot(qnorm(ps, 0, 1/sqrt(10) + 1), small_sample_median, main=title)
qqline(small_sample_median, col = 2)
title <- paste0("large")
qqplot(qnorm(ps, 0, 1/sqrt(1000) + 1), large_sample_median, main=title)
qqline(large_sample_median, col = 2)
```

## 2 Set up

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/babies.txt"
filename <- basename(url)
download.file(url, destfile=filename)
babies <- read.table("babies.txt", header=TRUE)
bwt.nonsmoke <- filter(babies, smoke==0) %>% select(bwt) %>% unlist 
bwt.smoke <- filter(babies, smoke==1) %>% select(bwt) %>% unlist
```

## 2.1

ans: 0.053
```{r}
N=10
set.seed(1)
nonsmokers <- sample(bwt.nonsmoke , N)
smokers <- sample(bwt.smoke , N)
dat <- c(smokers,nonsmokers)
obs <- mean(smokers) - mean(nonsmokers)

permutation <- function() {
    
  shuffle <- sample( dat  )
    smokersstar <- shuffle[1:N]
    nonsmokersstar <- shuffle[(N+1):(2*N)]
      return (mean(smokersstar)-mean(nonsmokersstar))

}
obs_permutation <- replicate(1000, permutation())
p_value <- mean(abs(obs_permutation) > abs(obs))
round(p_value, 4)
```

## 2.2

ans: 0.017
```{r}
N=10
set.seed(1)
nonsmokers <- sample(bwt.nonsmoke , N)
smokers <- sample(bwt.smoke , N)
obs <- median(smokers) - median(nonsmokers)

permutation <- function() {
    dat <- c(smokers,nonsmokers)
  shuffle <- sample( dat  )
    smokersstar <- shuffle[1:N]
    nonsmokersstar <- shuffle[(N+1):(2*N)]
      return (median(smokersstar) - median(nonsmokersstar))

}
obs_permutation <- replicate(1000, permutation())
p_value <- mean(abs(obs_permutation) > abs(obs))
round(p_value, 4)
```

## 3

ans: 6.513
```{r}
N=10
set.seed(1)
nonsmokers <- sample(bwt.nonsmoke , N)
smokers <- sample(bwt.smoke , N)
obs <- sd(smokers) - sd(nonsmokers)

set.seed(1)
bootstrap <- function(){
    nonsmokers <- sample(bwt.nonsmoke , N, replace = TRUE)
  smokers <- sample(bwt.smoke , N, replace = TRUE)
    return (sd(smokers) - sd(nonsmokers))

}
bootstrap_sd <- replicate(1000, bootstrap())
round(sd(bootstrap_sd), 4)
```
