---
title: 'Lecture 14: Clustering'
output: html_document
date: "2016/12/14"
---

# Clustering

## RMD_example 14.1: Read 照片分群 (pictcluster.csv)

```{r}
# Get the file url for the data
fileurl = "http://ghuang.stat.nctu.edu.tw/course/statmethods16/files/data/pictcluster.csv"
# Import the csv file into R
#   Note: stringsAsFactors=TRUE will screw up conversion to numeric!
pictcluster<-read.csv(fileurl, header=TRUE, stringsAsFactors=FALSE)
#   Note: read.csv treats the 1st column of pictcluster as the data, not the rownames, we have to correct this 
#         Get the 1st column as the rownames
names1<-pictcluster[,1]
#         Remove the 1st column from pictcluster
pictcluster<-pictcluster[,-1]
#         Assign the rownames
rownames(pictcluster)<-names1
colnames(pictcluster)
dim(pictcluster)

# 將上傳時間做轉換，以"天數"為單位，並相較於最早上傳的時間，計算出每筆資料與最早上傳的時間差
pictcluster[,c("上傳時間")]<-pictcluster[,c("上傳時間")]*10^8
t0<-as.POSIXlt(as.numeric(min(pictcluster[,c("上傳時間")])), origin="1970-01-01", tz="GMT")
tt<-as.POSIXlt(as.numeric(pictcluster[,c("上傳時間")]), origin="1970-01-01", tz="GMT")
pictcluster[,c("上傳時間")]<-as.numeric(difftime(tt, t0, unit="day"))

head(pictcluster)
```

## RMD_example 14.2: Agglomerative hierarchical clustering

```{r}
# Distance matrix between the rows of a data matrix
da<-dist(pictcluster, method="euclidean")
# Perform agglomerative hierarchical clustering with complete linkage
fita<-hclust(da, method="complete")
# Display dendogram (with labels at the same level)
plot(fita, label=rownames(pictcluster), hang=-1)
# Cut tree into 4 clusters
cutree(fita, k=4)
# Draw dendogram with red borders around the 4 clusters
rect.hclust(fita, k=4, border="red")
```

## RMD_example 14.3: Divisive hierarchical clustering

```{r}
# Install package "cluster" first, then load it into R
library(cluster)
# Distance matrix between the rows of a data matrix
dd<-dist(pictcluster, method="euclidean")
# Perform divisive hierarchical clustering
fitd<-diana(dd, diss=T)
# Display dendogram
plot(fitd)
# Cut tree into 4 clusters
cutree(fitd, k=4)
# Draw dendogram with red borders around the 4 clusters
rect.hclust(fitd, k=4, border="red")
```

## RMD_example 14.4: Heatmap

```{r}
# No dendrogram
heatmap(as.matrix(pictcluster), Rowv=NA, Colv=NA,
        xlab="變數", ylab="相片ID",
        cexRow=1, cexCol=1)
# With dendrogram
heatmap(as.matrix(pictcluster),
        xlab="變數", ylab="相片ID",
        cexRow=1, cexCol=1)
# With different color schemes
#   Install package "RColorBrewer" first, then load it into R
library(RColorBrewer)

heatmap(as.matrix(pictcluster), col=brewer.pal(9, "BuGn"),
        xlab="變數", ylab="相片ID",
        cexRow=1, cexCol=1)

# Use heatmap.2 and self-set dendrogram
#   Install packages "graphics", "gplots" and "RColorBrewer" first, then load them into R
library(graphics)
library(gplots)
library(RColorBrewer)

hcpict <- hclust(dist(pictcluster), method="complete")
dendpict <- as.dendrogram(hcpict)
hcvar <- hclust(as.dist(1-cor(pictcluster)), method="complete")
dendvar <- as.dendrogram(hcvar)

#   Create a blue -> purple colour palette
jBuPuFun <- colorRampPalette(brewer.pal(n = 9, "BuPu"))
paletteSize <- 256
jBuPuPalette <- jBuPuFun(paletteSize)

heatmap.2(as.matrix(pictcluster), Rowv=dendpict, Colv=dendvar, col=jBuPuPalette,
          xlab="變數", ylab="相片ID", scale="none", trace="none",
          cexRow=1, cexCol=1, density.info="none")
```

## RMD_example 14.5: K-means clustering 

```{r}
# 4 cluster solution
fitk<-kmeans(pictcluster, centers=4)
# Clustering results
fitk$cluster
# Get cluster means
aggregate(pictcluster, by=list(fitk$cluster), FUN=mean)
```

## RMD_example 14.6: K-medoids clustering 

```{r}
# Install package "cluster" first, then load it into R
library(cluster)
# Distance matrix between the rows of a data matrix
dp<-dist(pictcluster, method="euclidean")
# 4 cluster solution
fitp<-pam(dp, k=4, diss=T)
# Clustering results
fitp$cluster
# Get cluster means
aggregate(pictcluster, by=list(fitp$cluster), FUN=mean)
```

## RMD_example 14.7: Determine number of clusters

```{r}
wss<-(nrow(pictcluster)-1)*sum(apply(pictcluster,2,var))
for (i in 2:15) wss[i]<-sum(kmeans(pictcluster, centers=i)$withinss)
plot(1:15, wss, type="b", xlab="number of clusters",
     ylab="within clusters sum of squares")
```

## RMD_example 14.8: Silhouette plot

Silhouette plot can be used to assess cluster fit. Fisrt let do the silhouette plot for agglomerative hierarchical clustering.

```{r}
# Install package "cluster" first, then load it into R
library(cluster)

# Agglomerative hierarchical clustering
da<-dist(pictcluster, method="euclidean")
fita<-hclust(da, method="complete")
cuta<-cutree(fita, k=4)

# Silhouette plot
sia<-silhouette(cuta, da)
plot(sia)
#   with cluster-wise coloring
plot(sia, col=c("red", "green", "blue", "purple"))
```

Draw the silhouette plot for divisive hierarchical clustering.

```{r}
# Divisive hierarchical clustering
dd<-dist(pictcluster, method="euclidean")
fitd<-diana(dd, diss=T)
cutd<-cutree(fitd, k=4)

# Silhouette plot
sid<-silhouette(cutd, dd)
plot(sid, col=c("red", "green", "blue", "purple"))
```

Draw the silhouette plot for k-means clustering.

```{r}
# K-means clustering
dk<-dist(pictcluster, method="euclidean")
fitk<-kmeans(pictcluster, centers=4)

# Silhouette plot
sik<-silhouette(fitk$cluster, dk)
plot(sik, col=c("red", "green", "blue", "purple"))
```

Draw the silhouette plot for k-medoids clustering.

```{r}
# K-medoids clustering
dp<-dist(pictcluster, method="euclidean")
fitp<-pam(dp, k=4, diss=T)

# Silhouette plot
sip<-silhouette(fitp, dp)
plot(sip, col=c("red", "green", "blue", "purple"))
```

# Multidimensional scaling

## RMD_example 14.9: Multidimensional scaling

```{r}
# Euclidean distances between the rows
dM<-dist(pictcluster, method="euclidean")
# MDS
fitM<-cmdscale(dM, eig=T, k=2) # k is the number of dim
fitM # view results

# Plot solution 
x <- fitM$points[,1]
y <- fitM$points[,2]
plot(x, y, xlab="Coordinate 1", ylab="Coordinate 2", 
     main="Metric MDS",	type="n")
text(x, y, labels=row.names(pictcluster), cex=1)
```

