---
title: 'Lecture 15: Classification'
output: html_document
date: "2016/12/21"
---

# Classification

## RMD_example 15.1: Read 部落格分類 data

```{r}
# Get the file url for training data
fileurl = 'http://ghuang.stat.nctu.edu.tw/course/statmethods16/files/data/blogtrain.csv'
# Import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
blogtrain<-read.csv(fileurl, header=TRUE, stringsAsFactors=FALSE)
#   note: read.csv treats the 1st column of blogtrain.csv as the data, not the rownames,
#         we have to correct this
#         get the 1st column as the rownames
names1<-blogtrain[,1]
#         remove the 1st column from blogtrain
blogtrain<-blogtrain[,-1]
#         assign the rownames
rownames(blogtrain)<-names1
colnames(blogtrain)
dim(blogtrain)
head(blogtrain)

# Get the file url for testing data
fileurl = 'http://ghuang.stat.nctu.edu.tw/course/statmethods16/files/data/blogtest.csv'
# Import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
blogtest<-read.csv(fileurl, header=TRUE, stringsAsFactors=FALSE)
#   note: read.csv treats the 1st column of blogtest.csv as the data, not the rownames,
#         we have to correct this
#         get the 1st column as the rownames
names2<-blogtest[,1]
#         remove the 1st column from blogtest
blogtest<-blogtest[,-1]
#         assign the rownames
rownames(blogtest)<-names2
colnames(blogtest)
dim(blogtest)
head(blogtest)
```

## RMD_example 15.2: Linear discriminant analysis (LDA)

We firt "train" the model using `blogtrain`.

```{r}
# Install package "MASS" first, then load it into R
library(MASS)

# Create model formula with all variables as predictors
xname<-colnames(blogtrain)[-5]
(fmla <- as.formula(paste("as.factor(cateid) ~ ", paste(xname, collapse= "+"))))
fitlda<-lda(fmla, prior=c(1/3,1/3,1/3), data=blogtrain, na.action="na.omit")

# Show the results
fitlda
```

We can plot each observation in the space of the first 2 linear discriminants using the following codes. Points are identified with the class label.

```{r}
plot(fitlda)
```

A nice way of displaying the results of a LDA is to make a stacked histogram of the values of the discriminant function for the objects from different classes.

We can do this using the `ldahist()` function in R. For example, to make a stacked histogram of the 1st discriminant function’s values for objects of the three different `catei` classes, we type:

```{r}
lda.pred <- predict(fitlda)
ldahist(lda.pred$x[,1], g=blogtrain$cateid, type="both")
```

A stacked histogram of the 2nd discriminant function’s values for objects of the three different `catei` classes:

```{r}
ldahist(lda.pred$x[,2], g=blogtrain$cateid, type="both")
```

The `partimat( )` function in the `klaR` package can display the results of a linear classifications 2 variables at a time.

```{r}
# Install package "klaR" first, then load it into R
library(klaR)
partimat(fmla, data=blogtrain, method="lda")
```

Assess the accuracy of the prediction from leave-one-out cross-validation

```{r}
fitldac<-lda(fmla, prior=c(1/3,1/3,1/3), data=blogtrain, na.action="na.omit", CV=TRUE)
ct<-table(blogtrain$cateid, fitldac$class)
ct

# Total percent correct from leave-one-out cross-validation
sum(diag(prop.table(ct)))
```

Then, assess the accuracy of the prediction using the independent testing set `blogtest`

```{r}
fitlda<-lda(fmla, prior=c(1/3,1/3,1/3), data=blogtrain, na.action="na.omit")
btest<-blogtest[,-5]
testcl<-predict(fitlda, btest)$class
testcl
cttl<-table(blogtest$cateid, testcl)
cttl

# Total percent correct from the testing set
sum(diag(prop.table(cttl)))
```

## RMD_example 15.3: Quadratic discriminant analysis (QDA)

"Train" the model using `blogtrain`.

```{r}
# Create model formula with all variables as predictors
xname<-colnames(blogtrain)[-5]
(fmla <- as.formula(paste("as.factor(cateid) ~ ", paste(xname, collapse= "+"))))
fitqda<-qda(fmla, prior=c(1/3,1/3,1/3), data=blogtrain, na.action="na.omit")

# Show the results
fitqda

# Display the results of a quadratic classifications 2 variables at a time.
#   Install package "klaR" first, then load it into R
library(klaR)
partimat(fmla, data=blogtrain, method="qda")

# Assess the accuracy of the prediction from leave-one-out cross-validation
fitqdac<-qda(fmla, prior=c(1/3,1/3,1/3), data=blogtrain, na.action="na.omit", CV=TRUE)
ct<-table(blogtrain$cateid, fitqdac$class)
ct

# Total percent correct from leave-one-out cross-validation
sum(diag(prop.table(ct)))
```

Assess the accuracy of the prediction using the independent testing set `blogtest`

```{r}
fitqda<-qda(fmla, prior=c(1/3,1/3,1/3), data=blogtrain, na.action="na.omit")
btest<-blogtest[,-5]
testcq<-predict(fitqda, btest)$class
testcq
cttq<-table(blogtest$cateid, testcq)
cttq

# Total percent correct from the testing set
sum(diag(prop.table(cttq)))
```

