---
title: 'Lecture 8: Statistical decision making: hypothesis testing'
output: html_document
date: "2016/11/2"
---

# Hypothesis testing for continuous random variables

## RMD_example 8.1: Read the Gene expression microarray data (samplexprs.csv) 

```{r}
# Get the file url
fileurl <- "http://ghuang.stat.nctu.edu.tw/course/statmethods16/files/data/samplexprs.csv"

# Import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
samplexprs <- read.csv(fileurl, header=TRUE, stringsAsFactors=FALSE)

dim(samplexprs)
head(samplexprs[,1:10])
```

## RMD_example 8.2: One-sample t-test

```{r}
# Summary statistics for exprs of gene J00129
#   Install package "psych" first, then load it into R
library(psych)
J00129<-samplexprs[,'J00129']
describe(J00129)

# One-sample t-test
t.test(J00129, mu=-0.5) # Ho: mu=-0.5
```

## RMD_example 8.3: Two-sample t-test

```{r}
# summary statistics of J00129 by metastases
metastases<-samplexprs[,'metastases']
describeBy(J00129, metastases)

# two-sample t-test for J00129 over metastases
t.test(J00129~metastases, var.equal=T, na.action='na.omit')
t.test(J00129~metastases, var.equal=F, na.action='na.omit')
var.test(J00129~metastases, na.action='na.omit')
```

## RMD_example 8.4: ANOVA

```{r}
# summary statistics of J00129 by ERp
ERp<-samplexprs[,'ERp']
describeBy(J00129, ERp)

# more than two samples test for J00129 over ERp
anova(lm(J00129~as.factor(ERp)))
```

## RMD_example 8.5: Paired t-test

```{r}
# summary statistics for exprs of gene Contig29982_RC
Contig29982_RC<-samplexprs[,'Contig29982_RC']
describe(Contig29982_RC)

# paired t-test for comparing J00129 vs Contig29982_RC
t.test(J00129, Contig29982_RC, paired=T, na.action='na.omit')
```
# Measures of association and hypothesis testing for categorical data

## RMD_example 8.6: Binomial test

```{r}
# generating frequency tables
#   install package "gmodels", "epiR" first, then load them into R
library(gmodels)
library(epiR)

#   create ERs=1 if ERp>10, 0 o.w.
ERs<-rep(0, length(ERp))
ERs[!is.na(ERp) & ERp>10]<-1
ERs[is.na(ERp)]<-NA

# test for a single proportion
TAB<-xtabs(~ERs)
TAB
#   notice: p=Pr(ERs=0)
chisq.test(TAB) # Ho: p=0.5
chisq.test(TAB, p=c(0.4, 0.6)) # Ho: p=0.4
binom.test(TAB, p=0.5) # Ho: p=0.5
binom.test(TAB, p=0.4) # Ho: p=0.4
```

## RMD_example 8.7: Cohort and case-control studies

For exploring the data, we first create the $2 \times 2$ table.

```{r}
TAB1<-xtabs(~ERs+metastases)
TAB1
```

If the data are from a chort study, we can have relative risk and odds ratio. In the function `epi.2by2`, notice that "Exposed +" is for "ERs=0" and "Exposed -" is for "ERs=1", and "Outcome +" is for "metastases=0" and "Outcome -" is for "metastases=1"

```{r}
epi.2by2(TAB1, method='cohort.count')
```

If the data are from a case-control study, we can have the odds ratio.

```{r}
epi.2by2(TAB1, method='case.control')
```

## RMD_example 8.8: Chi2 / Fisher's exact test

```{r}
# chi2 / Fisher's exact test
CrossTable(ERs, metastases, prop.chisq=F, chisq=T, fisher=T)
```

## RMD_example 8.9: McNemar's test for matched-pair studies

```{r}
# good prognosis (1 to 34) matches poor prognosis (1 to 34)
Pg<-ERs[metastases==1][1:34]
Gg<-ERs[metastases==0][1:34]
TAB2<-xtabs(~Pg+Gg)
TAB2

# McNemar's test for ERs on prognosis
mcnemar.test(TAB2)
```

## RMD_example 8.10: Measure of agreement and its test

```{r}
# measure of agreement
#    on expression of J00129 and Contig29982_RC
Jexpry<-rep(0, length(J00129))
Jexpry[!is.na(J00129) & (J00129 > -0.5)]<-1
Jexpry[is.na(J00129)]<-NA
Cexpry<-rep(0, length(Contig29982_RC))
Cexpry[!is.na(Contig29982_RC) & (Contig29982_RC > -0.5)]<-1
Cexpry[is.na(Contig29982_RC)]<-NA
TAB3<-xtabs(~Jexpry+Cexpry)
TAB3

# kappa coefficient/test
#   install the package "fmsb" first, then load it into R
library(fmsb)
Kappa.test(TAB3)
```

## RMD_example 8.11: Mantel-Haenszel test

```{r}
# create age40=1 if age>40, 0 o.w.
age<-samplexprs[,'age']
age40<-rep(0, length(age))
age40[!is.na(age) & age>40]<-1
age40[is.na(age)]<-NA

TAB4<-xtabs(~ERs+metastases+age40)
TAB4
epi.2by2(TAB4, method='cohort.count')
mantelhaen.test(TAB4)
```

# Nonparametric statistical methods

## RMD_example 8.12: Nonparametric tests

```{r}
# sign test
#   install the package "BSDA" first, then load it into R
library(BSDA)
SIGN.test(J00129, md=-0.5) # Ho: mu=-0.5

# Wilcoxon signed-rank test
#   install the package "MASS" first, then load it into R
library(MASS)
wilcox.test(J00129, Contig29982_RC, paired=T)

# Wilcoxon rank-sum (Mann-Whitney) test
wilcox.test(J00129~metastases)

# Kruskal-Wallis test
kruskal.test(J00129~as.factor(ERp))
```
